---
output:
  html_document:
    theme: "flatly"
    toc: true
    toc_depth: 3
    toc_float: true
    keep_md: true
  pdf_document: default
---
Wine Data Exploration by Terry Verner
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(tidyverse)
library(ggcorrplot)
library(GGally)
library(ggthemes)
theme_update(panel.grid = element_blank())
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}

# Load the Data
raw_whites <- read_csv('./wineQualityWhites.csv')
white <- raw_whites[, 2:13]

raw_reds <- read_csv('./wineQualityReds.csv')
red <- raw_reds[, 2:13]


#raw_whites$type <- "white"
#raw_reds$type <- "red"
#wine <- bind_rows(raw_whites, raw_reds)
#wine$type <- as.factor(wine$type)

```


The dataset contains nearly 5000 records of white wine flavor profiles and
quality ratings.

# Univariate Plots Section

```{r echo=FALSE, warning=FALSE, Shape_of_the_Dataset}

dim(white)

```

The dataset includes 4898 observations of 12 features.

```{r echo=FALSE, warning=FALSE, Structure_of_Data}

str(white)

```

The data types for the features are 11 numeric (float) and 1 integer.

```{r echo=FALSE, warning=FALSE, Summary_Statistics}

summary(white)

```

Let's dive in to some univariate plotting to get a better look at a handful 
of these features and consider engineering new features.

```{r echo=FALSE, warning=FALSE, Table_Quality}

table(white$quality)

```

```{r echo=FALSE, warning=FALSE, Histogram_of_Quality}

ggplot(white, aes(x = quality)) +
    geom_histogram(binwidth = 0.5,
                   fill = 'white',
                   color = 'black') +
    scale_x_continuous(breaks = seq(0, 10, 
                                    by = 1.0),
                       limits = c(0, 10.0)) +
    geom_vline(xintercept = mean(white$quality, na.rm = TRUE),
               color = 'orange',
               size = 1,
               linetype = 5) +
    geom_vline(xintercept = median(white$quality, na.rm = TRUE),
               color = 'blue',
               size = 1,
               linetype = 3)

```

The distribution of quality is approximately normal. With a median of 6 and 
mean of 5.878, we can see the center of the data is slightly skewed toward the
higher values. I would say the data exhibits a very slight left skew.

```{r echo=FALSE, warning=FALSE, Alcohol_Histogram}

ggplot(white, aes(x = alcohol)) +
    geom_histogram(binwidth = 0.20, 
                   color = 'black',
                   fill = 'white') +
    scale_x_continuous(breaks = seq(min(white$alcohol), 
                                    max(white$alcohol), 
                                    by = 0.20)) +
    theme(axis.text.x=element_text(angle=90)) +
    geom_vline(xintercept = mean(white$alcohol, na.rm = TRUE),
               color = 'orange',
               size = 1,
               linetype = 5) +
    geom_vline(xintercept = median(white$alcohol, na.rm = TRUE),
               color = 'blue',
               size = 1,
               linetype = 3)

```

This is interesting because the data for alcohol is not normally distributed.
The data is right skewed with the mean and median values at 10.51 and 10.4 
respectively. However, the largest bin is 9.4 percent alcohol by volume.

```{r echo=FALSE, warning=FALSE, Boxplot_of_Alcohol_by_Quality}

ggplot(white, aes(x = as.factor(quality), y = alcohol)) +
    geom_boxplot() +
    xlab("quality")

```

It appears that in most, but not all cases, the alcohol content is more 
positively correlated with wine quality. The Pearson correlation coefficient is
`r cor(white$quality, white$alcohol)`.

```{r echo=FALSE, warning=FALSE, Volatile_Acidity_Histogram}

ggplot(white, aes(x = volatile.acidity)) +
    geom_histogram(binwidth = 0.025, 
                   color = 'black',
                   fill = 'white') +
    scale_x_continuous(breaks = seq(min(white$volatile.acidity), 
                                    max(white$volatile.acidity), 
                                    by = 0.025)) +
    theme(axis.text.x=element_text(angle=90)) +
    geom_vline(xintercept = mean(white$volatile.acidity, na.rm = TRUE),
               color = 'orange',
               size = 1,
               linetype = 5) +
    geom_vline(xintercept = median(white$volatile.acidity, na.rm = TRUE),
               color = 'blue',
               size = 1,
               linetype = 3)

```

The volatile acidity can, in high amounts, impart vinegar-like taste. The data
for this feature is significantly right skewed. I was hoping to use this 
feature to correlate with quality, but the few samples of high values might 
make that a futile endeavor.

```{r echo=FALSE, warning=FALSE, Boxplot_of_Volatile.acidity_by_Quality}

ggplot(white, aes(x = as.factor(quality), y = volatile.acidity)) +
    geom_boxplot() +
    xlab("quality")

```

The variability of volatile acidity across quality is not great and the 
correlation seems very weak at best. At least initially, this appears to be a
dead end from an analytical perspective.

```{r echo=FALSE, warning=FALSE, Citric_Acid_Histogram}

ggplot(white, aes(x = citric.acid)) +
    geom_histogram(binwidth = 0.05, 
                   color = 'black',
                   fill = 'white') +
    scale_x_continuous(breaks = seq(min(white$citric.acid), 
                                    max(white$citric.acid), 
                                    by = 0.1)) +
    geom_vline(xintercept = mean(white$citric.acid, na.rm = TRUE),
               color = 'orange',
               size = 1,
               linetype = 5) +
    geom_vline(xintercept = median(white$citric.acid, na.rm = TRUE),
               color = 'blue',
               size = 1,
               linetype = 3)

```

```{r echo=FALSE, warning=FALSE, Citric_Acid_Histogram_V2}

ggplot(white, aes(x = citric.acid)) +
    geom_histogram(binwidth = 0.05, 
                   color = 'black',
                   fill = 'white') +
    scale_x_continuous(breaks = seq(min(white$citric.acid), 
                                    max(white$citric.acid), 
                                    by = 0.05),
                       limits = c(min(white$citric.acid),
                                  quantile(white$citric.acid, 0.95))) +
    theme(axis.text.x=element_text(angle=90)) +
    geom_vline(xintercept = mean(white$citric.acid, na.rm = TRUE),
               color = 'orange',
               size = 1,
               linetype = 5) +
    geom_vline(xintercept = median(white$citric.acid, na.rm = TRUE),
               color = 'blue',
               size = 1,
               linetype = 3)

```

After zooming in to exclude outliers, the distribution is relatively normal. I
might be able to combine this feature with another, or others, and glean some 
significant insights from it.

```{r echo=FALSE, warning=FALSE, Boxplot_of_Citric.acid_by_Quality}

ggplot(white, aes(x = as.factor(quality), y = citric.acid)) +
    geom_boxplot() +
    xlab("quality")

```

Citric acid adds a freshness or light flavor to wine. The data for this
feature is right skewed. Similar to volatile acidity, I was hoping to use 
this feature to correlate with quality, but the few samples of high values 
might discount that effort as well.

```{r echo=FALSE, warning=FALSE, Free_Sulfur_Dioxide_Histogram}

ggplot(white, aes(x = free.sulfur.dioxide)) +
    geom_histogram(binwidth = 3, 
                   color = 'black',
                   fill = 'white') +
    scale_x_continuous(breaks = seq(min(white$free.sulfur.dioxide), 
                                    max(white$free.sulfur.dioxide), 
                                    by = 3),
                       limits = c(min(white$free.sulfur.dioxide),
                                  max(white$free.sulfur.dioxide))) +
    theme(axis.text.x=element_text(angle=90)) +
    coord_cartesian(xlim = c(min(white$free.sulfur.dioxide),
                             quantile(white$free.sulfur.dioxide, 0.95))) +
    geom_vline(xintercept = mean(white$free.sulfur.dioxide, na.rm = TRUE),
               color = 'orange',
               size = 1,
               linetype = 5) +
    geom_vline(xintercept = median(white$free.sulfur.dioxide, na.rm = TRUE),
               color = 'blue',
               size = 1,
               linetype = 3)

```

After zooming in to exclude outliers, the distribution of free sulfur dioxide
is normal. This is a feature I may be able to work with during later analysis.

```{r echo=FALSE, warning=FALSE, Boxplot_of_Free.sulfur.dioxide_by_Quality}

ggplot(white, aes(x = as.factor(quality), y = free.sulfur.dioxide)) +
    geom_boxplot() +
    xlab("quality")

```

```{r echo=FALSE, warning=FALSE, Boxplot_of_Free.sulfur.dioxide_by_Quality_V2}

ggplot(white, aes(x = as.factor(quality), y = free.sulfur.dioxide)) +
    geom_boxplot() +
    xlab("quality") +
    coord_cartesian(ylim = c(0, 100))

```

The free.sulfur.dioxide by quality, however, is not very significant in terms
of correlation. I will need to use this feature in conjunction with another.

```{r echo=FALSE, warning=FALSE, Create_New_Feature}

white$so2.percent <- white$free.sulfur.dioxide / white$total.sulfur.dioxide

```

Let's create a new feature as the percentage of free sulfur dioxide. This is
calculated by dividing `free.sulfur.dioxide` by `total.sulfur.dioxide`. The new
feature will be called so2.percent.

```{r echo=FALSE, warning=FALSE, SO2_Percentage_Histogram}

ggplot(white, aes(x = so2.percent)) +
    geom_histogram(fill = 'white',
                   color = 'black',
                   bins = 50) +
    xlab("Free / Total Sulfur Dioxide") +
    geom_vline(xintercept = mean(white$so2.percent, na.rm = TRUE),
               color = 'orange',
               size = 1,
               linetype = 5) +
    geom_vline(xintercept = median(white$so2.percent, na.rm = TRUE),
               color = 'blue',
               size = 1,
               linetype = 3)

```

I think this percentage feature can allow me greater insights into the quality
of white wines. SO2 is required to prevent oxidation, however, a balance must 
be found to prevent a strong sulfuric smell in the nose or the taste.

```{r echo=FALSE, warning=FALSE, Boxplot_of_so2_percent_by_Quality}

ggplot(white, aes(x = as.factor(quality), y = so2.percent)) +
    geom_boxplot() +
    xlab("quality")

```

# Univariate Analysis

#### What is the structure of your dataset?

The data is now structured with 12 continuous features and 1 discrete, categorical
output feature. With nearly 5000 observations, I may be able to glean more 
detailed insights or create more granular plots as compared to the red wines
dataset of 1600 observations.

#### What is/are the main feature(s) of interest in your dataset?

The primary feature I am interested in is the output variable, quality. If
the wine does not taste good, no one will drink it or even give much thought
to its characteristics influencing that quality.

#### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The other features I am interested in pertain to specific influences of taste.
These are residual sugar and various measures of acidity.

#### Did you create any new variables from existing variables in the dataset?

I created a single new variable as free sulfur dioxide divided by the total 
sulfur dioxide. This proporation is normally distributed and may allow a better
spread of the data while trying to extract meaningful insights.

#### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

While I did not have to clean the data or make it tidy, I did observe unusual
distributions of a handful of features. The unusual distribution could lead to
some interesting insights, or perhaps lead me to a dead end.

# Bivariate Plots Section

```{r echo=FALSE, warning=FALSE, Correlation_Matrix}

corr_matrix <- round(cor(white), 3)
corr_matrix

```

```{r echo=FALSE, warning=FALSE, Plot_Correlation_Matrix}

ggcorrplot(corr_matrix, method = 'square',
           colors = c('dark orange', 'light grey', 'dark blue'))

```

This correlation plot is not encouraging in determining very many strong 
relationships, at least at a rudimentary level. I will likely need to slice and 
manipulate my plots further to glean actionable insights. The multivariate 
plots should help drill down better in the section following.

```{r echo=FALSE, warning=FALSE, Alcohol_by_Quality}

ggplot(white, aes(quality, alcohol)) +
    geom_point(position = position_jitter(),
               alpha = 0.25)
    
```

Although faint, there may a positive correlation between quality and alcohol
content. The correlation coefficient of these two features is 
`r cor(white$quality, white$alcohol)`.

```{r echo=FALSE, warning=FALSE, Density_by_Alcohol}

ggplot(white, aes(alcohol, residual.sugar)) +
    geom_point(position = position_jitter(),
               alpha = 0.25) +
    coord_cartesian(ylim = c(0, 35))

```

This plot suggests a weak to moderate negative correlation between alcohol and
residual sugar. Higher residual sugar levels more frequently correspond to 
lower alcohol levels. This surprises me to certain degree because other types 
of strong alcohols try to make their beverage more palatable and to mask the 
strength of the alcohol. This, however, suggests many of the strong wines have
lower levels of residual sugar.

```{r echo=FALSE, warning=FALSE, pH_by_Fixed.Acidity}

ggplot(white, aes(fixed.acidity, pH)) +
    geom_point(position = position_jitter(),
               alpha = 0.25)

```

pH and fixed.acidity shows a moderate negative correlation as I expected. This
is due to acids being represented by lower values on the pH scale.

```{r echo=FALSE, warning=FALSE, Fixed.Acidity_by_Quality}

ggplot(white, aes(quality, fixed.acidity)) +
    geom_point(position = position_jitter(),
               alpha = 0.25)

```

I was hopeful the fixed acidity level might be well described by quality; 
however, the correlation appears to be weakly negative to not correlated.


```{r echo=FALSE, warning=FALSE, SO2_Percent_by_Quality}

ggplot(white, aes(quality, so2.percent)) +
    geom_point(alpha = 0.1)

```

The percentage of free sulfur dioxide compared to the total, did not end up as
descriptive as I anticipated it could be, nor predictive in terms of one being
described by the other.

```{r echo=FALSE, warning=FALSE, Residual.sugar_by_Density}

ggplot(white, aes(density, residual.sugar)) +
    geom_point(alpha = 0.1)

```

Even without zooming in to confirm at a smaller scale, I can see a strongly
positive correlation between density and residual sugars. This is expected
because sugar is typically heavier and more dense than the other chemicals
composing a beverage or other liquid solution.

```{r echo=FALSE, warning=FALSE, Residual.sugar_by_Density_V2}

ggplot(white, aes(density, residual.sugar)) +
    geom_point(alpha = 0.1) +
    coord_cartesian(xlim = c(min(white$density), 1.015),
                    ylim = c(min(white$residual.sugar), 40))

```

Zoomed in, the previous plot trendline is more easily observed. There is a 
correlation between density and residual.sugar

```{r echo=FALSE, warning=FALSE, so2.percent_by_sulphates}

ggplot(white, aes(sulphates, so2.percent)) +
    geom_point(position = position_jitter(),
               alpha = 0.25)

```

I was genuinely surprised there is effectively no correlation between sulphates
and so2.percent. I would expect one feature to influence the other, yet this
plot suggests that is not the case.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I think there may be relationships between multiple variables rather than
simply two variables. Most of the visualizations I developed did not 
demonstrate a strong or even moderate correlation. Only a handful of plots 
alone showed a relationship between the two variables.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I think the most interesting relationship I uncovered was the moderate,
negative correlation between alcohol and residual sugars. Other types of strong
alcohols add sugar and mask the strength of the alcohol. This, however, 
suggests many of the strong wines have lower levels of residual sugar.
There may be some process of wine creation that could explain this correlation.

### What was the strongest relationship you found?

The strongest relationship was one I expected: residual sugar vs. density. The
density of wine was strongly correlated with residual sugars and that is 
expected because any residual sugar would add density to a liquid solution.


# Multivariate Plots Section

```{r echo=FALSE, warning=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
                                                                         units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, warning=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, warning=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, warning=FALSE, Plot_Three}

```

### Description Three

------
  
  # Reflection
  
  > **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!